---
output:
  html_document:
    css: HAL.css
    self_contained: false
    toc: yes
    toc_float: yes
    code_folding: hide
date: "2026-10-8"
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'docs/index.html')) })
---

```{r libraries, include=FALSE, results='hide'}

library(glptools)

glp_load_packages(T)

library(lubridate)
library(readxl)
library(tidyr)
library(sf)
library(dplyr)
library(plotly)
```

# Housing Alliance of Louisville (HAL) Dashboard

## Homelessness:

-   Goal: Secure stable housing for 250 individuals experiencing chronic homelessness by 2027.

## Eviction:

-   Goal: Eliminate evictions for parents with children age 0-5 in Louisville
-   Plot of eviction filings, judgements, warrants:

```{r eviction, echo=FALSE, message=FALSE, warning=FALSE}
library(showtext)
library(sysfonts)

font_add_google("Poppins", "poppins")
showtext_auto()

load("data/eviction/eviction.RData")

eviction_summary_qtr <- eviction %>%
  st_drop_geometry() %>%
  filter(filing_date >= lubridate::ym("2021-01"),
         filing_date < lubridate::ym("2025-10")) %>%
  mutate(year = year(filing_date),
         quarter = quarter(filing_date)) %>%
  group_by(year, quarter) %>%
  summarise(
    filings = n(),
    judgements = sum(JPF),
    warrants = sum(warrant_flag)
  ) %>%
  mutate(period = paste0(year, " Q", quarter),
         qtr_num = year + (quarter - 1) / 4) %>%
  pivot_longer(
    cols = c(filings, judgements, warrants),
    values_to = "n",
    names_to = "Type"
  )

p <- ggplot(eviction_summary_qtr,
            aes(
              x = qtr_num,
              y = n,
              color = Type,
              group = Type,
              text = paste0(period, "<br>", Type, ": ", comma(n))
            )) +
  scale_color_manual(values = c("#F4354F", "#FF8699", "#FFD3D6")) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 1) +
  scale_x_continuous(breaks = 2021:2025, labels = 2021:2025) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = NULL) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  theme_bw()

p <- ggplotly(p, tooltip = "text") %>%
  layout(legend = list(title = list(text = "")))
p
```

## Affordable Housing:

-   Goal: Create or preserve 15,000 affordable housing units by 2027
-   Current count: 7016 units (My Louisville Home)
-   Data from OHCD coming this week (10/7)
-   Landscape:
    -   Map (NHPD, OHCD, LAHTF)
-   How has affordability changed?
    -   Percent cost burdened over time (ACS)
    -   Current renter burden (ACS) vs. new renter burden (Zillow)
    -   Other option
    
```{r zillow_map}
load("data/housing/zillow_hvi.Rdata")

z_hvi <- zillow_hvi %>%
  select(RegionName, everything()) %>%
  rename(zip = RegionName, value = 2)

zhvi_growth <- z_hvi %>%
  select(zip, X2000.01.31, X2026.08.31) %>%
  mutate(growth_pct = (X2026.08.31 / X2000.01.31 - 1) * 100) %>%
  select(zip, growth_pct)

map_zip <- glptools::map_zip

zhvi_growth_map <- map_zip %>%
  mutate(zip = as.integer(zip)) %>%
  left_join(zhvi_growth, by = "zip") %>%
  ggplot() +
  geom_sf(aes(fill = growth_pct),
          color = "white",
          linewidth = 0.2) +
  scale_fill_gradientn(
    colors = c("#FFD3D6", "#FF8699", "#F4354F"),
    labels = scales::label_percent(scale = 1),
    name = "Growth"
  ) +
  labs(title = "Home Value Growth, 2000–2026", subtitle = "Zillow Home Value Index") +
  theme_void() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

zhvi_growth_map
```

-   GROUPED BAR CHART?
    -   Affordability by income groups (TBD)
        -   Less than 20% income to housing
        -   20-29% income to housing
        -   30+% income to housing

### Louisville Mean Hourly Wage (2025): \$29.96 = \$62,316.80 per year

(U.S. Bureau of Labor Statistics)

### Louisville Minimum Wage: \$7.25 = \$15,080 per year

```{r}
df <- read_excel("data/housing/housing_cost_percentage.xlsx") %>%
  janitor::clean_names()

income_levels <- c("<20k", "20k to 34k", "35k to 49k", "50k to 74k", "75k+")

income_cb_renter <- df %>%
  filter(tenure == "Renter") %>%
  mutate(income = factor(income, levels = income_levels)) %>%
  pivot_longer(
    c(
      "less_than_20_percent",
      "between_20_29_percent",
      "over_30_percent"
    ),
    names_to = "% income spent on housing",
    values_to = "percent"
  ) %>%
  mutate(
    `% income spent on housing` = recode(
      `% income spent on housing`,
      less_than_20_percent = "<20%",
      between_20_29_percent = "20% to 29%",
      over_30_percent = "30%+"
    ),
  )

income_cb_owner <- df %>%
  filter(tenure == "Owner") %>%
  mutate(income = factor(income, levels = income_levels)) %>%
  pivot_longer(
    c(
      "less_than_20_percent",
      "between_20_29_percent",
      "over_30_percent"
    ),
    names_to = "% income spent on housing",
    values_to = "percent"
  ) %>%
  mutate(
    `% income spent on housing` = recode(
      `% income spent on housing`,
      less_than_20_percent = "<20%",
      between_20_29_percent = "20% to 29%",
      over_30_percent = "30%+"
    )
  )

library(showtext)
library(sysfonts)

font_add_google("Poppins", "Poppins")
showtext_auto()

plot_income_cb_renter <- ggplot(
  income_cb_renter,
  aes(
    x = income,
    y = percent,
    fill = `% income spent on housing`,
    group = `% income spent on housing`
  )
) +
  geom_col(position = position_dodge(width = 0.5), width = 0.5) +
  labs(title = "Cost Burden by Income Group (Renter)", x = "Income Group", y = "Percent") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("#FFD3D6", "#FF8699", "#F4354F")) +
  theme_minimal(base_family = "poppins", base_size = 22) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      family = "poppins"
    ),
    plot.title = element_text(
      face = "bold",
      size = 26,
      family = "poppins"
    ),
    legend.position = "top"
  )

plot_income_cb_renter

plot_income_cb_owner <- ggplot(
  income_cb_owner,
  aes(
    x = income,
    y = percent,
    fill = `% income spent on housing`,
    group = `% income spent on housing`
  )
) +
  geom_col(position = position_dodge(width = 0.5), width = 0.5) +
  labs(title = "Cost Burden by Income Group (Owner)", x = "Income Group", y = "Percent") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("#FFD3D6", "#FF8699", "#F4354F")) +
  theme_minimal(base_family = "poppins", base_size = 22) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      family = "poppins"
    ),
    plot.title = element_text(
      face = "bold",
      size = 26,
      family = "poppins"
    ),
    legend.position = "top"
  )

plot_income_cb_owner

```
